\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{microtype}
\usepackage[hidelinks]{hyperref}
\usepackage{url}

\newcommand{\podium}{\textsc{Podium}}
\newcommand{\dd}{\mathrm{d}}

\title{\podium: A Verification-Oriented Open-Source Library for\\
Rendezvous, Proximity Operations, and Docking GNC}
\author{Adi Oltean\thanks{\texttt{github.com/adi-oltean/podium}. The library and this
manuscript were developed with substantial AI assistance (Anthropic Claude);
all physics claims are enforced by the repository's test suite.}}
\date{July 2026 --- v0.1 preprint}

\begin{document}
\maketitle

\begin{abstract}
We present \podium{}, an open-source (MIT) Python library and simulation
sandbox for guidance, navigation, and control (GNC) of spacecraft rendezvous,
proximity operations, and docking (RPOD) in low and medium Earth orbit,
designed from the outset for translation to flight C code validated by
abstract interpretation. A survey of the open-source ecosystem finds no
maintained tool that combines validated relative-motion dynamics, RPOD
guidance primitives, deterministic simulation, and a credible path to
verified flight software; \podium{} targets exactly that gap. The library
enforces a strict two-layer architecture: a \emph{static subset} algorithm
core (pure step functions, fixed array shapes, compile-time loop bounds,
machine-readable range contracts on every scalar) mechanically translatable
to analyzable C, beneath a full-Python sandbox for truth models and analysis.
We describe the implemented v0.1 kernels---Clohessy--Wiltshire and
Yamanaka--Ankersen state transitions, a nonlinear dual-propagation truth
model with $J_2$ and differential drag, and pulsed minimum-impulse-bit
terminal docking control---together with the validation methodology:
independent cross-implementation checks, quadratic-scaling separation of
linearization error from implementation error, and CI-enforced validity
envelopes for the linear models. A case study validates the docking laws
end-to-end against an uncooperative black-box plant, SpaceX's public ISS
docking simulator, via browser automation: system identification recovered
an undocumented per-frame actuation model, a sign-inverted rate display, and
a frame-rate-dependent control gain, and the campaign surfaced two genuine
control defects---a pulse-deadband stall whose standing offset
$\tau\,\Delta v/2$ scales with frame rate, cured by sub-pulse dithering---
before the autopilot achieved repeatable docking success. We close with the
designed convex guidance stack (lossless convexification on relative-motion
STMs, successive convexification for six-degree-of-freedom docking) and the
planned contract-to-annotation verification pipeline.
\end{abstract}

\section{Introduction}

Autonomous rendezvous, proximity operations, and docking is moving from a
capability of a few national programs to a routine commercial requirement:
satellite servicing, active debris removal, orbital assembly, and crew and
cargo logistics all fly some version of the same terminal problem---bring a
chaser from kilometers away to millimeter-per-second contact inside tight
lateral, angular, and rate gates \cite{fehse2003}. The GNC algorithms
involved are mature in the literature, yet the \emph{software} situation is
lopsided. Flight-proven RPOD stacks are closed and proprietary. Open-source
astrodynamics is rich in propagators and orbit-determination machinery but
thin exactly where RPOD lives: validated relative-motion models, approach
corridors and keep-out zones, docking-phase control, passive abort safety,
and---most conspicuously---any deliberate route from prototype Python to
flight-grade, formally analyzable C.

\podium{} \cite{podium2026} is an attempt to fill that gap in the open. Its
premise is that verifiability is an architectural property, not a
post-processing step: if flight algorithms are written from day one in a
restricted \emph{static subset} of Python with contracts as data, the
eventual C translation is mechanical, and the resulting code lands in
precisely the dialect that industrial abstract interpreters are known to
prove runtime-error-free---the approach validated on the ESA Automated
Transfer Vehicle docking software by Astr\'ee in 2008 \cite{kastner2010}.

This preprint documents the v0.1 state of the library. Its contributions:

\begin{enumerate}
  \item a two-layer, verification-oriented architecture for GNC libraries,
        with machine-readable interval contracts on every core scalar
        (Section~\ref{sec:arch});
  \item implemented and cross-validated relative-motion kernels, including
        a Yamanaka--Ankersen state transition matrix, released under a
        permissive license together with a truth-model validation
        methodology and measured validity envelopes (Sections~\ref{sec:kernels}--\ref{sec:truth});
  \item pulsed (minimum-impulse-bit) terminal docking control laws with a
        closed-form deadband-stall analysis, near-zone gain scheduling, and
        sub-pulse dithering (Section~\ref{sec:docking});
  \item a black-box validation case study in which the docking laws fly
        SpaceX's public ISS docking simulator to repeatable success,
        including the system identification of its undocumented plant and
        two control defects the campaign surfaced (Section~\ref{sec:casestudy});
  \item a designed (not yet implemented) convex guidance stack and
        verification pipeline, stated as such (Section~\ref{sec:design}).
\end{enumerate}

\section{Related work and ecosystem gap}
\label{sec:related}

A July 2026 survey of open-source spacecraft simulation and GNC tooling
(archived in the repository) examined Basilisk \cite{schulte2020basilisk},
NASA's 42, Trick, cFS, and GMAT, ESA-adjacent Orekit and tudatpy, the
archived poliastro, nyx-space, and RPOD-specific efforts including AFRL's
CW-based reinforcement-learning environments and the MuJoCo-based
SmallSatSim. No maintained tool combines: (i) validated relative-motion
dynamics---notably, no surveyed tool ships an open Yamanaka--Ankersen
implementation; (ii) docking contact coupled to those dynamics; (iii)
RPOD-relevant sensor and actuator models; (iv) GNC algorithm hooks between
simulation and flight code; (v) first-class deterministic Monte Carlo; and
(vi) a path to verified flight C. Incumbent licenses are fractured
(ISC, NOSA, AGPL, ``government-purpose''), leaving permissive RPOD territory
essentially unoccupied. \podium{} deliberately does \emph{not} compete on
general simulation fidelity: Basilisk remains the recommended high-fidelity
backend, and Orekit/GMAT serve as cross-validation oracles.

On the verification side, \podium{} follows three precedents: the
Astr\'ee/ATV result that abstract interpretation can prove docking flight
code free of runtime errors when the code is one synchronous loop of pure,
statically-bounded step functions \cite{kastner2010}; Feron-style credible
autocoding, where controllers carry their invariants as machine-checkable
contracts \cite{feron2013}; and the embedded-solver code generation dialect
of CVXPYgen and its relatives---static allocation, fixed iteration counts,
\texttt{math.h} as the only dependency \cite{cvxpygen2022}. For closed-loop
verification, the ARCH-COMP spacecraft rendezvous benchmark
\cite{chan2017} defines the community-standard hybrid model that
\podium{}'s mode logic is designed to remain traceable to.

\section{Architecture: the two-layer rule}
\label{sec:arch}

Everything in \podium{} respects one boundary. The \textbf{flight core}
(\texttt{podium.core}, plus flight-side functions in \texttt{control} and
\texttt{guidance}) obeys the \emph{static subset}: pure step functions with
no globals, I/O, clock access, or randomness; fixed array shapes known at
import time; loops with compile-time constant bounds (solvers are
\emph{anytime}: fixed maximum iterations, status code, best iterate); no
recursion or function pointers; guarded divisions and square roots; and
saturation at every physical boundary. The \textbf{sandbox} (truth dynamics,
gain synthesis, Monte Carlo, plotting, solver prototyping) may use the full
scientific-Python stack. The simulation engine calls flight-core functions
through the same signature they will have after C translation.

Contracts are declared once, as data, and consumed four ways:
\begin{center}
\begin{verbatim}
@contract(n=Interval(1e-5, 1e-2), e=Interval(0.0, 0.9))
def stm(n, e, theta0, dt): ...
\end{verbatim}
\end{center}
(i) runtime checks during simulation; (ii) comment-annotation blocks
(\texttt{[range(lo,hi)]}, \texttt{[len(6)]}, \texttt{PROVE(...)}) for the
external abstract-interpretation tool at C translation; (iii) a generated
analysis harness that draws inputs nondeterministically from the declared
ranges---the analyzer's operating envelope, as in the ATV and A380
campaigns; (iv) API documentation. Unconstrained scalars are the leading
cause of failed interval-domain proofs, so the rule is absolute: every core
scalar parameter carries a range contract.

Determinism is likewise non-negotiable: fixed-step integration, one seeded
generator, no platform-dependent reductions---identical configuration and
seed give bit-identical trajectories, which makes every Monte Carlo outlier
exactly replayable.

\section{Relative-motion kernels}
\label{sec:kernels}

Frames and conventions are fixed library-wide: target-centered LVLH with
$x$ radial (zenith), $y$ along-track, $z$ cross-track; scalar-first
quaternions rotating body to reference; SI units throughout.

\subsection{Clohessy--Wiltshire}

For a circular target orbit with mean motion $n$, the classical equations
\cite{clohessy1960}
\begin{equation}
\ddot x = 3n^2x + 2n\dot y + u_x,\qquad
\ddot y = -2n\dot x + u_y,\qquad
\ddot z = -n^2 z + u_z
\end{equation}
admit the closed-form state transition matrix implemented in
\texttt{core.cw}, with two-impulse targeting by inversion of the
position--velocity block.

\subsection{Yamanaka--Ankersen state transition}

For an elliptic target ($0 \le e < 1$), \podium{} implements the
Yamanaka--Ankersen state transition \cite{yamanaka2002} for the
Tschauner--Hempel equations. With true anomaly $\theta$ as independent
variable, $\rho = 1 + e\cos\theta$, and the standard substitution
$\tilde{\mathbf r} = \rho\,\mathbf r$ with derivatives taken with respect to
$\theta$, the in-plane equations reduce to
\begin{equation}
\tilde x'' = \frac{3}{\rho}\,\tilde x + 2\tilde y', \qquad
\tilde y'' = -2\tilde x', \qquad
\tilde z'' = -\tilde z .
\end{equation}
Writing $s=\rho\sin\theta$, $c=\rho\cos\theta$, and
$J = k^2(t-t_0)$ with $k^2 = \sqrt{\mu/p^3}$, the implemented in-plane
fundamental matrix on $(\tilde x,\tilde y,\tilde x',\tilde y')$ is
\begin{equation}
\Phi_f(\theta, J) =
\begin{bmatrix}
s & c & 2-3esJ & 0\\[2pt]
c\,(1{+}1/\rho) & -s\,(1{+}1/\rho) & -3\rho^2 J & 1\\[2pt]
s' & c' & -3e\,(s'J + \sin\theta/\rho) & 0\\[2pt]
-2s & e-2c & -3\,(1-2esJ) & 0
\end{bmatrix},
\label{eq:yafund}
\end{equation}
each column verified by direct substitution into the Tschauner--Hempel
system during development. The physical-space transition over a time step
is assembled as
$\Phi = \Lambda^{-1}(\theta_1)\,\tilde\Phi(\theta_1,\theta_0)\,\Lambda(\theta_0)$,
where $\Lambda$ maps physical LVLH states to the tilde variables and the
anomaly propagation solves Kepler's equation with a fixed twenty-iteration
Newton recursion (the denominator $1-e\cos E \ge 1-e$ is bounded away from
zero by the eccentricity contract $e \le 0.9$)---bounded iteration being a
static-subset requirement, not a style choice.

\subsection{Validation methodology and validity envelopes}
\label{sec:envelopes}

Every physics kernel in \podium{} must carry a ``receipt'': a test against
an independent reference with a tolerance derived from physics rather than
tuned to pass. The YA kernel's receipts illustrate the methodology:

\begin{enumerate}
  \item \textbf{Independent truth.} The STM is compared against numerical
        integration of the \emph{nonlinear} elliptic relative dynamics
        (Section~\ref{sec:truth}) across $e \in \{0, 0.05, 0.15, 0.35\}$.
  \item \textbf{Error-scaling separation.} The residual against truth must
        scale \emph{quadratically} with separation (halving the initial
        offset must quarter the error): linearization error is
        $O(\lVert\delta\mathbf r\rVert^2/a)$, while any implementation
        defect scales linearly. This single test discriminates ``model
        error, as designed'' from ``bug.''
  \item \textbf{Structural identities.} Exact reduction to the CW STM at
        $e=0$; composition $\Phi(t_0{\to}t_2) =
        \Phi(t_1{\to}t_2)\Phi(t_0{\to}t_1)$; inverse via negative time
        steps; $\det\Phi = 1$.
  \item \textbf{CI-enforced envelopes.} The measured one-orbit position
        error obeys $\varepsilon \le C(e)\,\lVert\delta\mathbf r\rVert^2/a$
        with $C \approx 40$ for $e \le 0.05$ and $C \approx 200$ at
        $e = 0.2$ (the nonlinear term is amplified near perigee), asserted
        in CI together with the quadratic scaling so the bound cannot
        silently loosen. For context, at 1\,km separation and low
        eccentricity this is centimeter-level model error per orbit.
        The same harness documents \emph{why the YA kernel exists}: at
        1\,km separation the CW error exceeds the YA error by a factor
        ${>}30$ at $e=0.05$ and ${>}100$ at $e=0.2$.
\end{enumerate}

\section{Nonlinear truth model}
\label{sec:truth}

The truth layer (\texttt{dynamics.nonlinear}) makes no relative-motion
approximation: both spacecraft are integrated in ECI under identical force
models---two-body, $J_2$, and exponential-atmosphere drag with a co-rotating
atmosphere and per-spacecraft ballistic coefficients---and the relative
state is formed by differencing in the target's LVLH frame. Double-precision
differencing error at LEO radii ($\sim$nanometers) is far below truth-model
needs. The LVLH frame's angular velocity is treated exactly,
\begin{equation}
\boldsymbol\omega_{\mathrm{LVLH}} =
\Bigl[\; \frac{r}{h}\,(\mathbf a_{\mathrm{pert}}\cdot\hat{\mathbf z}),\;
0,\; \frac{h}{r^2} \;\Bigr],
\end{equation}
including the out-of-plane precession component that is often dropped;
a dedicated test central-differences the position history under active
$J_2$ and drag and requires agreement with the reported relative velocity
to better than $2\times10^{-5}$\,m/s. Further receipts: secular RAAN drift
matching $-\tfrac{3}{2}nJ_2(R_E/p)^2\cos i$ over ten orbits; drag
semi-major-axis decay matching the analytic per-orbit estimate;
differential-drag along-track drift with the correct sign; two-body energy
conservation to $10^{-10}$ relative at the CI step size. The YA and truth
implementations are fully independent, so they cross-check each other.

\section{Pulsed terminal docking control}
\label{sec:docking}

The docking phase of many vehicles---and, conveniently, the control model
of the simulator in Section~\ref{sec:casestudy}---is \emph{pulsed}: each
thruster firing imparts a fixed velocity increment $\Delta v$ (a minimum
impulse bit, MIB), and the controller chooses integer pulse counts.
\texttt{podium.control.docking} provides three laws:

\paragraph{Closing-rate profile.} $v_{\mathrm{cmd}}(r) =
\mathrm{clamp}(r/\tau_a,\, v_{\mathrm{contact}},\, v_{\mathrm{far}})$: an
exponential-decay glideslope in range with a firm, capture-safe contact
floor.

\paragraph{Gain-scheduled proportional nulling.} Lateral offsets and
attitude errors are driven by saturated proportional velocity/rate commands
$u = -\mathrm{clamp}(p/\tau,\pm u_{\max})$ with a shorter time constant
inside a near zone. The scheduling is not cosmetic. With pulse quantization
and a half-pulse deadband, a proportional law stalls once
$|p|/\tau < \Delta v/2$, leaving a standing offset
\begin{equation}
p_{\mathrm{stall}} = \tfrac{1}{2}\,\tau\,\Delta v .
\label{eq:stall}
\end{equation}
With far-field gains alone this evaluates to $0.6$\,m laterally and
$0.25^\circ$ per axis at the case study's pulse sizes---\emph{outside} the
$0.2$\,m/$0.2^\circ$ docking gates. The near-zone constants place the stall
at $0.1$\,m and $0.05^\circ$. \podium{}'s click-dynamics unit tests (a
per-axis integrator plant actuated only by fixed increments) caught exactly
this failure in the first implementation; the tests encode the sim's
success gates as assertions.

\paragraph{Pulse allocator.} $n = \mathrm{round}_0\!\bigl((v_{\mathrm{cmd}}
- v_{\mathrm{est}})/\Delta v\bigr)$ with a half-pulse deadband
(round-toward-zero) so the vehicle never chatters about its setpoint, and a
per-cycle saturation. For fine tracking the deadband is optionally replaced
by \emph{sub-pulse dithering}: the fractional pulse is carried into the
next control cycle (first-order $\Sigma\Delta$ modulation), so the average
applied rate tracks commands far below one pulse per cycle. Section
\ref{sec:casestudy} shows why this matters.

\section{Case study: black-box validation against the SpaceX ISS docking
simulator}
\label{sec:casestudy}

SpaceX's public ISS docking simulator \cite{isssim} implements Crew
Dragon's manual docking mode: pulsed translation and rotation inputs, HUD
telemetry, and the operative gates ($<0.2$\,m lateral, $<0.2^\circ$ per
axis, $<0.2$\,m/s closing at contact). As an \emph{uncooperative,
undocumented plant} with a real success criterion, it makes an unusually
honest validation target. A Playwright-driven harness injects the exact
autopilot script that human users paste into the browser console; the
autopilot reads the sim's telemetry and calls the sim's own control
handlers (identical to clicking its buttons).

\subsection{System identification}

Because nothing about the plant is documented, the control mapping was
recovered empirically (button-probe campaigns sampling raw telemetry) and
by reading the minified source. Findings, all load-bearing:
(i) translation state is meters \emph{per frame}, applied along
\emph{capsule body} axes, one \texttt{translationPulseSize} per click, with
the displayed $(x,y,z)$ mapping to body $(-z,+x,+y)$ and displayed lateral
values being the \emph{negative} of the port's body coordinates;
(ii) the attitude ``rate'' display is a click counter whose sign is
\emph{inverted} relative to the true error drift---the camera's angular
velocity is proportional to \emph{minus} the accumulated click register,
with a frame-rate-dependent gain;
(iii) the docking contact test is a raycast whose reach shrinks to
$0.1$\,m at terminal range, so an approach faster than $\sim0.1$\,m/frame
\emph{tunnels through the port geometrically}---observed directly: an
otherwise perfect approach at $0.159$\,m/frame passed through the station
untouched;
(iv) the success gate requires each world-frame velocity component below
$0.004$\,m/frame with all three angles below $0.2^\circ$ at contact.

The flight autopilot therefore counts \texttt{requestAnimationFrame} to
convert its SI control laws to frame units at the actuation boundary,
measures the attitude plant gain at engagement with a five-pulse probe
(re-estimated in flight), and creeps the final meters at $0.003$\,m/frame.

\subsection{Failure modes found by test}

The campaign surfaced three defects, each with a lesson:

\begin{enumerate}
  \item \textbf{Polarity inversion} (all three rotation axes): assumed
        button directions were backwards, producing positive feedback and
        an uncontrollable spin in the first live run. In-flight
        sign-learning proved noise-unstable; the cure was offline system
        identification and hard-coded, verified polarities.
  \item \textbf{Lateral limit cycle into structure}: aggressive lateral
        capture near the station produced $\pm 6$\,m oscillations at
        one-cycle latency and a grazing contact with station structure at
        $\sim$8\,m range. Cured by distance-scheduled lateral velocity
        caps and a 10\%-half-angle approach corridor that pauses closing
        whenever the corridor is violated, with a gate-hold inside 4\,m.
  \item \textbf{Frame-rate-dependent deadband stall}: after headless
        verification passed end-to-end, a live 60\,fps run docked
        $0.2$--$0.3$\,m below the port center and failed. In frame units
        the stall radius of Eq.~\eqref{eq:stall} becomes
        $p_{\mathrm{stall}} = \tfrac{1}{2}\tau\,\Delta v_{\mathrm{click}} f$:
        $0.24$\,m at $f{=}60$\,fps but only $0.02$\,m at the harness's
        SwiftShader $5$\,fps---the bug was \emph{invisible at every frame
        rate the automated tests ran at}. Sub-pulse dithering
        (Section~\ref{sec:docking}) removes the dead zone at any frame
        rate; regression and live runs then both dock cleanly.
\end{enumerate}

After the fixes, the autopilot achieves the simulator's official SUCCESS
outcome repeatably: headless (harness-verified, with wall-clock compression
by state teleports between independently verified flight phases) and on a
60\,fps desktop from the standard 200\,m start with randomized initial
attitude, in roughly three minutes of flight time. Beyond the anecdote, the
methodological point is that \emph{cross-fidelity validation---unit-level
click dynamics, slow headless end-to-end, and full-rate live runs---caught
disjoint bug sets}; no single tier would have sufficed.

\section{Designed: convex guidance and the verification pipeline}
\label{sec:design}

Two major subsystems are designed and documented in the repository but not
yet implemented; we state them here for completeness and to invite
scrutiny.

\paragraph{Layered convex guidance.} Layer~0 poses far/mid-field planning
as one-shot convex programs on the exact CW/YA discretizations: LP/SOCP/QP
costs; approach cones as second-order-cone constraints; rotating-hyperplane
keep-out zones; Breger--How passive-safety constraints \cite{breger2008};
and thrust-annulus lossless convexification following the G-FOLD lineage
\cite{acikmese2007,acikmese2011}, whose rendezvous exactness was
established by Lu and Liu \cite{lu2013}---with validity checks
(controllability, normality, discrete-time repair) shipped as part of the
implementation rather than left to the user, since keep-out zones are not
losslessly convexifiable \cite{malyuta2022} and interval-active state
constraints weaken the guarantees. Layer~1 adds successive convexification
for six-degree-of-freedom docking: penalized-trust-region iteration
\cite{szmuk2018} as the default engine (flight pedigree via NASA SPLICE
\cite{splice2025}), SCvx* \cite{oguri2023} where a feasibility guarantee is
required, and state-triggered constraints for plume and minimum-impulse-bit
logic \cite{malyuta2020docking}. The prototype-to-embedded path is
CVXPY with disciplined parameterized programming, generating
static-memory, fixed-iteration C solvers \cite{cvxpygen2022}---the same
dialect the verification pipeline expects, with onboard feasibility
anchored by \cite{dueri2017,weiss2015}.

\paragraph{Verification pipeline.} The static subset plus contracts are
designed so that a small AST-walking emitter can translate the core to C
with contracts rendered as analyzer annotations and \texttt{prove()} calls
as proof obligations, validated externally by an abstract-interpretation
tool in the Astr\'ee tradition \cite{kastner2010}; golden-vector
equivalence between the Python and C artifacts validates the translation
itself. Closed-loop safety is planned as reachability regression on the
ARCH-COMP rendezvous benchmark \cite{chan2017}, keeping the code-level mode
machine traceable to the verified hybrid model, in the spirit of
proof-carrying control \cite{feron2013}.

\section{Conclusion}

\podium{} v0.1 demonstrates that a GNC library can be simultaneously
open, physically rigorous, and verification-oriented without exotic
tooling: the discipline is architectural. The implemented kernels ship with
receipts---independent truth models, scaling-law bug discriminators, and
CI-enforced validity envelopes---and the docking control stack has been
validated end-to-end against an adversarially undocumented external plant,
an exercise that surfaced exactly the class of unit-convention and
quantization bugs that destroy real rendezvous missions. The near-term
roadmap is the sim engine and Monte Carlo harness, the Layer-0 convex
guidance stack, and the C emitter that turns the contract discipline into
proofs. Contributions are welcome; every physics claim in the repository is
gated by the test suite, and the authors would particularly value
independent replication of the validity envelopes and adversarial review of
the static-subset rules.

\bibliographystyle{unsrt}
\bibliography{references}

\end{document}
